---
title: "PHB228 Homework1"
author: "Hongjun Chen"
format: html
editor: visual
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(bench)
library(palmerpenguins)
library(purrr)
```

## Part 1: Vectorization and Data Structures

### 1.1 Vectorization

#### (a)

```{r}
vectorized_zscore <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

loop_zscore <- function(x) {
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  z <- numeric(length(x))
  
  for (i in seq_along(x)) {
    z[i] <- (x[i] - m) / s
  }
  
  z
}

x_iris <- iris$Sepal.Length

head(vectorized_zscore(x_iris))
head(loop_zscore(x_iris))
all.equal(vectorized_zscore(x_iris), loop_zscore(x_iris))
```

From the output, both functions give the same z-scores. The vectorized version applies the formula to the whole vector at once, while the loop computes it one value at a time. Even though the results are the same, the implementation style is different.

#### (b)

```{r}
bench_results <- bench::mark(
  vectorized = vectorized_zscore(x_iris),
  loop = loop_zscore(x_iris),
  iterations = 50,
  check = FALSE
)

bench_results

median_times <- as.numeric(bench_results$median)
names(median_times) <- bench_results$expression

barplot(
  median_times,
  main = "Median Execution Time",
  ylab = "Time (seconds)",
  xlab = "Method"
)
```

From the benchmark results, the vectorized method is slightly faster than the loop. This makes sense because vectorized operations in R are optimized internally and avoid the overhead of explicit loops.

#### (c)

```{r}
chick_growth <- ChickWeight %>%
  group_by(Chick) %>%
  arrange(Time, .by_group = TRUE) %>%
  mutate(
    Growth.Rate = (weight - lag(weight)) / (Time - lag(Time)),
    Weight.Gain.Percent = (last(weight) - first(weight)) / first(weight) * 100
  ) %>%
  ungroup()

chick_summary <- chick_growth %>%
  group_by(Diet, Chick) %>%
  summarize(
    avg_growth_rate_chick = mean(Growth.Rate, na.rm = TRUE),
    max_weight_chick = max(weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Diet) %>%
  summarize(
    average_growth_rate = mean(avg_growth_rate_chick, na.rm = TRUE),
    mean_maximum_weight = mean(max_weight_chick, na.rm = TRUE),
    .groups = "drop"
  )

head(chick_growth)
chick_summary
```

Growth.Rate shows how quickly each chick\'s weight changes between time points. Looking at the summary table, we can compare how different diets affect both growth rate and maximum weight.

### 1.2 Data Structures and Functional Programming

#### (a)

```{r}
data(penguins, package = "palmerpenguins")

species_names <- unique(na.omit(penguins$species))

penguin_list <- lapply(species_names, function(sp) {
  dat <- penguins[penguins$species == sp & !is.na(penguins$species), ]
  attr(dat, "sample_size") <- nrow(dat)
  dat
})

names(penguin_list) <- species_names

lapply(penguin_list, function(x) attr(x, "sample_size"))
```

Each element of the list corresponds to one species. I also stored the sample size as an attribute so it\'s easy to check how many observations each group has.

#### (b)

```{r}
numeric_penguins <- penguins %>%
  select(where(is.numeric))

means_lapply <- lapply(numeric_penguins, mean, na.rm = TRUE)
means_map_dbl <- map_dbl(numeric_penguins, mean, na.rm = TRUE)

means_lapply
means_map_dbl
```

`lapply()` returns a list, while `map_dbl()` returns a numeric vector. The numeric vector is easier to use directly in later calculations.

#### (c)

```{r}
mean_mass_by_species <- tapply(
  penguins$body_mass_g,
  penguins$species,
  mean,
  na.rm = TRUE
)

mean_mass_by_species_sex <- tapply(
  penguins$body_mass_g,
  list(penguins$species, penguins$sex),
  mean,
  na.rm = TRUE
)

mean_mass_by_species
mean_mass_by_species_sex
```

The first result gives the average body mass for each species. The second one breaks it down further by both species and sex, so we can see more detailed differences.

#### (d)

```{r}
v <- c(1, 2, 3, 4, 5)
v_ref <- v

cat("Before modification:\n")
print(v)
print(v_ref)
cat("Identical before modification:", identical(v, v_ref), "\n")

v_ref[3] <- 99

cat("\nAfter modification:\n")
print(v)
print(v_ref)
cat("Identical after modification:", identical(v, v_ref), "\n")
```

At first, both variables refer to the same object. After modifying one of them, R creates a copy instead of changing the original. This is why only v_ref changes while v stays the same.

#### (e)

```{r}
inefficient_code <- function() {
  result <- numeric(0)
  for (i in 1:10000) {
    result <- c(result, i^2)
  }
  result
}

preallocated_code <- function() {
  result <- numeric(10000)
  for (i in 1:10000) {
    result[i] <- i^2
  }
  result
}

time_inefficient <- system.time(inefficient_result <- inefficient_code())
time_preallocated <- system.time(preallocated_result <- preallocated_code())

time_inefficient
time_preallocated
all.equal(inefficient_result, preallocated_result)
```

The preallocated version runs faster because the memory is allocated once at the beginning. In contrast, repeatedly using c() forces R to keep creating new copies, which slows things down.

## Part 2: Matrix Operations and Decompositions

### 2.1 Matrix Operations

#### (a)

```{r}
A <- matrix(c(4, 2, 1, 5), nrow = 2)
B <- matrix(c(3, 1, 0, 2), nrow = 2)
C <- matrix(c(1, 3, 5, 2, 4, 6), nrow = 2)

A
B
C

A_plus_B <- A + B
A_elementwise_B <- A * B
A_matrix_B <- A %*% B
tA_matrix_C <- t(A) %*% C
A_inverse <- solve(A)

A_plus_B
A_elementwise_B
A_matrix_B
tA_matrix_C
A_inverse
```

`A + B` adds corresponding entries of the two matrices. `A * B` performs elementwise multiplication, while `A %*% B` performs matrix multiplication. The expression `t(A) %*% C` first transposes `A` and then multiplies it by `C`. Finally, `solve(A)` computes the inverse of `A`.

#### (b)

```{r}
is_symmetric <- function(M) {
  if (!is.matrix(M)) {
    stop("Input must be a matrix.")
  }
  if (nrow(M) != ncol(M)) {
    stop("Input matrix must be square.")
  }
  isTRUE(all.equal(M, t(M)))
}

is_symmetric(A)
is_symmetric(B)
is_symmetric(t(A) %*% A)
```

A matrix is symmetric if it is square and equal to its transpose. Both `A` and `B` are not symmetric, but `t(A) %*% A` is symmetric.

### 2.2 Eigenvalue Decomposition

#### (a)

```{r}
M <- matrix(c(4, 2, 2, 3), nrow = 2)

eig_M <- eigen(M)

eig_M$values
eig_M$vectors
```

The eigen() function gives the eigenvalues and eigenvectors of M. I use these results in the next step to check whether M %*% x equals lambda* x.

#### (b)

```{r}
for (i in seq_along(eig_M$values)) {
  lambda <- eig_M$values[i]
  x_vec <- eig_M$vectors[, i]
  
  left_side <- M %*% x_vec
  right_side <- lambda * x_vec
  
  cat("Eigenpair", i, "\n")
  print(left_side)
  print(right_side)
  print(all.equal(as.numeric(left_side), as.numeric(right_side)))
}
```

For each eigenvalue-eigenvector pair, the equation `M %*% x = lambda * x` holds up to small numerical rounding error.

#### (c)

```{r}
P <- eig_M$vectors
D <- diag(eig_M$values)

M_reconstructed <- P %*% D %*% solve(P)

M
M_reconstructed
all.equal(M, M_reconstructed)
```

The reconstructed matrix matches the original matrix, which confirms that the eigenvalue decomposition `M = P D P^{-1}`.

### 2.3 SVD and QR Decomposition

#### (a) SVD

```{r}
X <- matrix(c(3, 2, 2, 2, 3, -2, 2, -2, 3), nrow = 3)

svd_X <- svd(X)

svd_X$d
svd_X$u
svd_X$v

rank1_X <- svd_X$d[1] * svd_X$u[, 1] %*% t(svd_X$v[, 1])

frobenius_error <- norm(X - rank1_X, type = "F")

X
rank1_X
frobenius_error
```

The rank-1 approximation keeps only the largest singular value. Since the smaller singular values are left out, the reconstruction is not exact and the Frobenius error is not zero.

#### (b) QR and Least Squares

```{r}
set.seed(228)
n <- 100
x <- runif(n, -5, 5)
y <- 3 + 2 * x + rnorm(n, sd = 1)

X_design <- cbind(1, x)

qr_X <- qr(X_design)
Q <- qr.Q(qr_X)
R <- qr.R(qr_X)

beta_qr <- solve(R) %*% t(Q) %*% y

lm_fit <- lm(y ~ x)

beta_qr
coef(lm_fit)
all.equal(as.numeric(beta_qr), as.numeric(coef(lm_fit)))
```

The QR solution gives the same least squares estimates as `lm()`, up to numerical rounding. QR decomposition is useful for least squares because it avoids explicitly forming `X^T X`, which improves numerical stability.

## Part 3: Optimization and Simulation

### 3.1 Gradient Descent

#### (a)

```{r}
rosenbrock <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  (1 - x1)^2 + 100 * (x2 - x1^2)^2
}

rosenbrock(c(-1.2, 1.0))
```

This evaluates the Rosenbrock function at a given point, which is commonly used as a test problem for optimization.

#### (b)

```{r}
rosenbrock_grad <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  
  grad_x1 <- -2 * (1 - x1) - 400 * x1 * (x2 - x1^2)
  grad_x2 <- 200 * (x2 - x1^2)
  
  c(grad_x1, grad_x2)
}

rosenbrock_grad(c(-1.2, 1.0))
```

The gradient tells us the direction of the steepest increase, so moving in the opposite direction helps minimize the function.

#### (c)

```{r}
gradient_descent <- function(init, grad_fn, learning_rate, max_iter, tol) {
  x <- init
  
  for (iter in 1:max_iter) {
    grad <- grad_fn(x)
    
    if (sqrt(sum(grad^2)) < tol) {
      return(list(
        solution = x,
        value = rosenbrock(x),
        iterations = iter,
        converged = TRUE
      ))
    }
    
    x <- x - learning_rate * grad
  }
  
  list(
    solution = x,
    value = rosenbrock(x),
    iterations = max_iter,
    converged = FALSE
  )
}
```

The algorithm updates the parameters step by step using the gradient. It stops when the gradient becomes very small or when it reaches the maximum number of iterations.

#### (d)

```{r}
gd_result <- gradient_descent(
  init = c(-1.2, 1.0),
  grad_fn = rosenbrock_grad,
  learning_rate = 0.001,
  max_iter = 10000,
  tol = 1e-6
)

gd_result
```

From the result, the algorithm moves toward the minimum near (1, 1). The output shows the final estimate and whether it converged.

### 3.2 Comparison with Built-in Methods

#### (a)

```{r}
bfgs_result <- optim(
  par = c(-1.2, 1.0),
  fn = rosenbrock,
  gr = rosenbrock_grad,
  method = "BFGS"
)

bfgs_result
```

BFGS converges faster because it uses gradient information and approximates curvature, making it more efficient than basic gradient descent.

#### (b)

```{r}
nelder_result <- optim(
  par = c(-1.2, 1.0),
  fn = rosenbrock,
  method = "Nelder-Mead"
)

method_comparison <- tibble(
  method = c("Gradient Descent", "BFGS", "Nelder-Mead"),
  x1 = c(gd_result$solution[1], bfgs_result$par[1], nelder_result$par[1]),
  x2 = c(gd_result$solution[2], bfgs_result$par[2], nelder_result$par[2]),
  function_value = c(gd_result$value, bfgs_result$value, nelder_result$value),
  iterations = c(gd_result$iterations, bfgs_result$counts["function"], nelder_result$counts["function"]),
  distance_to_true_min = c(
    sqrt(sum((gd_result$solution - c(1, 1))^2)),
    sqrt(sum((bfgs_result$par - c(1, 1))^2)),
    sqrt(sum((nelder_result$par - c(1, 1))^2))
  )
)

method_comparison
```

All methods aim to find the same minimum. BFGS usually performs best, while Nelder-Mead works without gradients but may take longer. Gradient descent is simpler but depends on the learning rate.

### 3.3 Simulation Study

#### (a)

```{r}
run_simulation <- function(n_samples, n_reps, dist_type) {
  results <- vector("list", n_reps)
  
  for (i in 1:n_reps) {
    if (dist_type == "normal") {
      sample_data <- rnorm(n_samples, mean = 0, sd = 1)
    } else if (dist_type == "t3") {
      sample_data <- rt(n_samples, df = 3)
    } else if (dist_type == "contaminated") {
      indicator <- rbinom(n_samples, size = 1, prob = 0.10)
      sample_data <- ifelse(
        indicator == 1,
        rnorm(n_samples, mean = 0, sd = 10),
        rnorm(n_samples, mean = 0, sd = 1)
      )
    } else {
      stop("Unknown distribution type.")
    }
    
    results[[i]] <- tibble(
      dist_type = dist_type,
      mean_est = mean(sample_data),
      median_est = median(sample_data),
      trimmed_mean_est = mean(sample_data, trim = 0.10)
    )
  }
  
  bind_rows(results)
}
```

This function simulates data repeatedly and computes different estimators so we can compare their behavior.

#### (b)

```{r}
set.seed(228)

sim_normal <- run_simulation(30, 1000, "normal")
sim_t3 <- run_simulation(30, 1000, "t3")
sim_contaminated <- run_simulation(30, 1000, "contaminated")

sim_all <- bind_rows(sim_normal, sim_t3, sim_contaminated)

sim_long <- sim_all %>%
  pivot_longer(
    cols = c(mean_est, median_est, trimmed_mean_est),
    names_to = "estimator",
    values_to = "estimate"
  )

simulation_summary <- sim_long %>%
  group_by(dist_type, estimator) %>%
  summarize(
    bias = mean(estimate) - 0,
    variance = var(estimate),
    mse = mean((estimate - 0)^2),
    .groups = "drop"
  )

simulation_summary
```

The true center is 0 for all three simulated distributions, so bias and MSE are calculated relative to 0.

#### (c)

```{r}
simulation_summary
```

The mean works well for normal data, but for heavy-tailed or contaminated data, the median and trimmed mean are more stable because they are less affected by outliers.

## Part 4: Bootstrap Methods

### 4.1 Bootstrap Confidence Intervals

#### (a)

```{r}
bootstrap_mean <- function(data, B = 2000, ci_type = "percentile", conf = 0.95) {
  n <- length(data)
  theta_hat <- mean(data)
  
  boot_means <- replicate(B, {
    boot_sample <- sample(data, size = n, replace = TRUE)
    mean(boot_sample)
  })
  
  alpha <- 1 - conf
  se <- sd(boot_means)
  
  if (ci_type == "percentile") {
    ci <- quantile(boot_means, probs = c(alpha / 2, 1 - alpha / 2))
  } else if (ci_type == "basic") {
    q <- quantile(boot_means, probs = c(1 - alpha / 2, alpha / 2))
    ci <- 2 * theta_hat - q
  } else if (ci_type == "normal") {
    ci <- c(theta_hat - qnorm(1 - alpha / 2) * se,
            theta_hat + qnorm(1 - alpha / 2) * se)
  } else {
    stop("Unknown CI type.")
  }
  
  list(
    estimate = theta_hat,
    se = se,
    ci_type = ci_type,
    lower = as.numeric(ci[1]),
    upper = as.numeric(ci[2]),
    boot_means = boot_means
  )
}

set.seed(42)
returns <- rnorm(50, mean = 0.08, sd = 0.12)
returns <- returns + rexp(50, rate = 20) - 0.05

set.seed(228)
boot_percentile <- bootstrap_mean(returns, B = 2000, ci_type = "percentile")
boot_basic <- bootstrap_mean(returns, B = 2000, ci_type = "basic")
boot_normal <- bootstrap_mean(returns, B = 2000, ci_type = "normal")

bootstrap_ci_table <- tibble(
  method = c("percentile", "basic", "normal"),
  estimate = c(boot_percentile$estimate, boot_basic$estimate, boot_normal$estimate),
  se = c(boot_percentile$se, boot_basic$se, boot_normal$se),
  lower = c(boot_percentile$lower, boot_basic$lower, boot_normal$lower),
  upper = c(boot_percentile$upper, boot_basic$upper, boot_normal$upper)
)

bootstrap_ci_table
```

The table reports the bootstrap standard error and the 95% confidence interval for the mean using percentile, basic, and normal bootstrap methods.

#### (b)

```{r}
hist(
  boot_percentile$boot_means,
  breaks = 30,
  main = "Bootstrap Distribution of the Mean",
  xlab = "Bootstrap Mean"
)

abline(v = c(boot_percentile$lower, boot_percentile$upper), lty = 2, lwd = 2)
abline(v = c(boot_basic$lower, boot_basic$upper), lty = 3, lwd = 2)
abline(v = c(boot_normal$lower, boot_normal$upper), lty = 4, lwd = 2)

legend(
  "topright",
  legend = c("Percentile CI", "Basic CI", "Normal CI"),
  lty = c(2, 3, 4),
  lwd = 2,
  bty = "n"
)
```

The bootstrap distribution looks slightly skewed, but the three intervals are very close. I would use the percentile interval here because it follows the empirical bootstrap distribution directly and does not require a normality assumption.

### 4.2 Bootstrap Hypothesis Testing

#### (a)

```{r}
bootstrap_ttest <- function(control, treatment, B = 2000) {
  observed_diff <- mean(treatment) - mean(control)
  
  pooled_mean <- mean(c(control, treatment))
  control_null <- control - mean(control) + pooled_mean
  treatment_null <- treatment - mean(treatment) + pooled_mean
  
  boot_diffs <- replicate(B, {
    boot_control <- sample(control_null, size = length(control), replace = TRUE)
    boot_treatment <- sample(treatment_null, size = length(treatment), replace = TRUE)
    mean(boot_treatment) - mean(boot_control)
  })
  
  p_value <- mean(abs(boot_diffs) >= abs(observed_diff))
  
  list(
    observed_diff = observed_diff,
    boot_diffs = boot_diffs,
    p_value = p_value
  )
}

set.seed(123)
control <- rnorm(40, mean = 10, sd = 2)
treatment <- rnorm(40, mean = 11.5, sd = 2)

set.seed(228)
boot_test <- bootstrap_ttest(control, treatment, B = 2000)

boot_test$observed_diff
boot_test$p_value

hist(
  boot_test$boot_diffs,
  breaks = 30,
  main = "Bootstrap Null Distribution",
  xlab = "Difference in Means Under Null"
)

abline(v = boot_test$observed_diff, lwd = 2, lty = 2)
abline(v = -boot_test$observed_diff, lwd = 2, lty = 2)
legend(
  "topright",
  legend = "Observed difference",
  lty = 2,
  lwd = 2,
  bty = "n"
)
```

The bootstrap test estimates the null distribution of the difference in means by resampling after centering the two groups to have the same mean. The p-value is calculated as the proportion of bootstrap null differences at least as extreme as the observed difference.

#### (b)

```{r}
standard_ttest <- t.test(treatment, control)

standard_ttest
boot_test$p_value
```

Both methods lead to similar conclusions here. The t-test relies on assumptions, while the bootstrap method is more flexible and does not depend on normality as strongly.

## Resources Used

I used course materials and R documentation for functions such as bench::mark(), eigen(), svd(), qr(), and optim().
